Honors Thesis Proposal\\Speculative Execution in Particle Swarm Optimization

### 2009-10-02 honors_thesis/proposal/paper.tex
Proposal is finished

\title{Honors Thesis Proposal}
\author{Matthew Gardner}
\date{\today}

@@

I will study the feasibility and potential performance gain of a technique
which we call speculative execution in particle swarm optimization.  By
speculative execution I mean the use of additional computational resources,
presumably extra processors running in parallel, to perform two or more
iterations of particle swarm optimization at a time, speeding up the algorithm
significantly.

\section{Procedures}

The bulk of the work for this project will be writing code and running
experiments.  I have already read many conference papers on particle swarm
optimization to know the current state of the field, so I will know what the
results need to be compared to.  In implementing speculative execution, I will
take care that the algorithm behaves exactly like the original, standard
particle swarm optimization (PSO) algorithm in every aspect except the
speed-up.  My work can then be accurately compared to other work in the field.

In running experiments, many trials will be run and averaged, as there is a
random component in this algorithm.  I will use statistical tests for all
comparisons to the original PSO algorithm.  The success criterion will be a
better performance of the algorithm as measured by the lowest function
evaluation seen over a number of function evaluations for several benchmark and
real-world problems.  Speculative execution requires more function evaluations 
per particle than a na\"ive parallel PSO, %TODO
PSO with the same number of particles, so the two algorithms will be compared
at points where they use equal amounts of function evaluations per iteration.

@@

\subsection{Speculative Execution}

@@

\begin{tabular}{|l|l|}
\hline
Run preliminary tests to ensure feasibility & February 28, 2009 \\
\hline
Finish writing and debugging code & April 30, 2009 \\
\hline
Finish gathering data and statistics & June 30, 2009 \\
\hline
Complete analysis of data and begin writing thesis & September 1, 2009 \\
\hline
Submit completed copy of thesis to advisor & November 15, 2009 \\
\hline
Submit paper detailing results to CEC 2010 & December 20, 2009 \\
\hline
Final thesis and portfolio submitted & February 1, 2010 \\
\hline
Thesis defense completed & March 1, 2010 \\
\hline
Final four thesis copies submitted & March 8, 2010 \\
\hline
\end{tabular}

>>>>

\title{Honors Thesis Proposal\\Speculative Execution in Particle Swarm
Optimization}
\author{Matthew Gardner}
\date{\today}

@@

I will study the feasibility and potential performance gain of a technique
which we call speculative execution in particle swarm optimization.  By
speculative execution I mean the use of additional computational resources,
presumably extra processors running in parallel, to perform two or more
iterations of particle swarm optimization at a time, thus finding better values
faster.

\section{Procedures}
\label{sec:proc}

\subsection{Speculative Execution}
\label{sec:specex}

\subsection{Planned Experiments}

I will implement speculative execution in particle swarm optimization.  I will
take care that my implementation can behave exactly like the original particle
swarm optimization, so a fair comparison can be made.  At each iteration a 
number of positions are speculatively evaluated, and to make it behave like
the original PSO all that needs to be done is throw away the particles that
do not match the actual execution path of the original particles.

After showing that speculative execution reproduces PSO without affecting
results, I will expand the idea to use the information gathered from
speculative particles.  The positions that are speculatively evaluated are
potentially, and probably provably, worthwhile samples from the sampling
distribution of PSO.  Throwing them away discards useful information that could
help the algorithm in more ways than just speeding it up.  The end result would
be an algorithm that not only performs twice as many iterations as standard
PSO, but takes useful samples and potentially does better while going faster.

I will test this algorithm on standard benchmark functions, including at least
the Sphere, Rastrigin and Griewank functions.  I will compare the performance
of the speculative algorithms to the standard PSO on those functions with
varying swarm sizes.

\subsection{Evaluation}

In running experiments, I will run many trials and average the results, as
there is a random component in this algorithm.  I will use statistical tests,
such as a paired permutation test, for all comparisons to the original PSO
algorithm.  The success criterion will be a better performance of the algorithm
as measured by the lowest function evaluation seen over a number of function
evaluations.  Performing speculative execution in PSO requires more function
evaluations per particle than regular PSO, so to be fair the algorithms will be
compared with configurations where each uses the same number of function
evaluations per iteration, not the same number of particles.

@@

Before going forward with this idea, I ran some basic, preliminary tests that
show that for some functions, performing more iterations of PSO with fewer
particles does better than performing fewer iterations with more particles.
The tests were very preliminary but were enough to establish the feasibility of
using speculative execution to improve the performance of PSO on some
functions.

Since then I have finished coding both speculative algorithms mentioned in
Section~\ref{sec:proc}.  Further, though still preliminary, testing shows again
that these methods perform better than standard PSO on a number of functions.
What remains to be done is to do more exhaustive tests, collecting results that
can be published.  I anticipate that the results that I find as I continue
testing these algorithms will drive further improvements, but the direction of
those improvements has yet to be seen.

@@

\begin{tabular}{|l|l|}
\hline
Run preliminary tests to ensure feasibility & February 28, 2009 \\
\hline
Finish writing and debugging code & April 25, 2009 \\
\hline
Finish gathering data and statistics & August 31, 2009 \\
\hline
Complete analysis of data and begin writing thesis & October 1, 2009 \\
\hline
Submit completed copy of thesis to advisor & November 15, 2009 \\
\hline
Submit paper detailing results to CEC 2010 & December 20, 2009 \\
\hline
Final thesis and portfolio submitted & February 1, 2010 \\
\hline
Thesis defense completed & March 1, 2010 \\
\hline
Final four thesis copies submitted & March 8, 2010 \\
\hline
\end{tabular}
